---
title: "Exploratory Data Analysis"
author: "Russel Luber"
date: "2025-11-04"
output: html_document
---

```{r}
library(tidyverse)
library(here)
```

Load in the data.

```{r}
fights <- read_csv(here("data/clean/fight_data.csv"))
```


```{r}
fighters <- read_csv(here("data/raw/fighters_data_raw.csv"))
```


# Data Validation

```{r}
glimpse(fighters)
```

```{r}
fighters %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "column",
    values_to = "na_count"
  ) %>%
  arrange(desc(na_count))
```

```{r}
unique(fighters$stance)
```

```{r}
fighters %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(
    n = sum(!is.na(value)),
    missing = sum(is.na(value)),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    q25 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    q75 = quantile(value, 0.75, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    .groups = "drop"
  )
```



Snapshot of the columns, col types, and values.
```{r}
glimpse(fights)
```

## String Clean Up

Remove leading and trailing whitespace in entries from columns that are strings (characters).

```{r}
fights <- fights %>%
  mutate(across(where(is.character), ~ str_trim(.x, side="both")))
```


## Duplicate Rows

I built a fighter-centric dataset, which means the key is `(fighter_id, fight_id)`: one row per fighter per fight.

Any duplicate rows?

```{r}
fights %>% count(fighter_id, fight_id) %>% filter(n > 1)
```

No duplicate rows. Good.

## Check the Dates

According to [Wikipedia] (https://en.wikipedia.org/wiki/Ultimate_Fighting_Championship), the UFC debuted its first event on November 12, 1993. Sanity check: there should be no fights in the dataset before that!

```{r}
fights %>%
  summarise(
    earliest_fight = min(date),
    latest_fight = max(date)
  )
```

The earliest fight in the dataset happened on March 11, 1994. Great.


Just in case `summarise()` somehow missed anything:
```{r}
fights %>%
  filter(date < as.Date("1993-11-12") | date > Sys.Date()) %>%
  arrange(date)
```

Good. No fight data before the UFC was founded and no future fight data that hasn't happened yet.


How many unique fights are there (that we have data on)?

```{r}
fights %>%
  summarise(n_unique_fights = n_distinct(fight_id))
```

8407 fights. Since there are two fighters involved in every match, we have 16814 rows in our `d` dataframe in total.

Are there missing values (`NANs`) in the data?

```{r}
fights %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "column",
    values_to = "na_count"
  ) %>%
  arrange(desc(na_count))
```

No missing data across all entries of the 47 columns.
- Need to review fetching and parsing scripts to ensure no missing data at this stage is a good thing.
- For each of the columns, did I replace `NANs` with something else? Like a 0 maybe?

## Verify Results

Check that there are only four outcomes in each row of the `res` (result) column: Win (W), Loss (L), Draw (D), or No Contest (NC). 

```{r}
fights %>%
  distinct(res) %>%
  arrange(desc(res))
```

Good. There are only four outcomes, as expected.

In a fight, there are only 3 possible combinations of outcomes:

1. One fighter wins, the other loses.
2. The fight is ruled a draw, in which case both fighters get a "D" on their record.
3. The fight is a no contest, in which case both fighters get an "NC" on their record.

```{r}
res_patterns <- fights %>%
  count(fight_id, res) %>%
  arrange(fight_id, res) %>%
  group_by(fight_id) %>%
  summarise(
    n_rows  = sum(n),
    pattern = paste0(res, ":", n, collapse = ","),
    .groups = "drop"
  ) %>%
  mutate(is_valid = n_rows == 2 & pattern %in% c("D:2", "NC:2", "L:1,W:1", "W:1,L:1"))

invalid_ids <- res_patterns %>% filter(!is_valid) %>% pull(fight_id)

offenders   <- fights %>% filter(fight_id %in% invalid_ids)
```


```{r}
summarise(offenders)
```

```{r}
# Check that ALL fights (unique fight_id) have valid entries for the res column
fights %>% distinct(fight_id) %>% count() %>% pull(n) %>% as.integer() == sum(res_patterns$is_valid)
```

No fight had two winners or two losers or only one no contest (or draw). Every fight has a valid entry for the `res` column.

## Weight Classes

The UFC has 8 official weight classes for men: 

1. Flyweight
2. Bantamweight
3. Featherweight
4. Lightweight
5. Welterweight
6. Middleweight
7. Light Heavyweight
8. Heavyweight


For women, the UFC has 4 official weight classes:
1. Strawweight
2. Flyweight
3. Bantamweight
4. Featherweight


And there's an additional weight class called "Catchweight", which is an agreed-upon weight limit for a fight that falls outside of the standard, officially-recognized weight classes.
- A fight is usually in catchweight if one or both of the fighters "missed weight."

```{r}
fights %>%
  distinct(gender, weight_class) %>%
  arrange(gender, weight_class)
```


```{r}
# In the UFC, the Strawweight division has only ever been for women. No men's strawweight division. Verify that
fights %>% filter(gender == "Men" & weight_class == "Strawweight")
```

How many unique fights are there across `(gender, weight_class)` combinations?

```{r}
fights %>% 
  distinct(fight_id, gender, weight_class) %>% 
  count(gender, weight_class, sort = TRUE)
```

Set weight classes for men and women as factors.

```{r}
levels_all <- c("Strawweight", "Flyweight","Bantamweight","Featherweight","Lightweight",
  "Welterweight","Middleweight","Light Heavyweight","Heavyweight","Catchweight")
```

```{r}
fights <- fights %>%
  mutate(weight_class = factor(weight_class, levels = levels_all))
```


Set gender as factor.

```{r}
gender_levels = c("Men", "Women")
```


```{r}
fights <- fights %>%
  mutate(gender = factor(gender, levels = gender_levels))
```


## Count Variables Cannot Be Negative

Variables like `strikes_landed`, `ctrl_time_s`, `td_stuffed`, and many (in fact most) of the variables in the dataset can be 0 but can't be negative.

```{r}
summary_nonneg <- fights %>%
  summarise(across(where(is.numeric), ~ sum(. < 0, na.rm = TRUE))) %>%
  tidyr::pivot_longer(everything(), names_to = "column", values_to = "n_negatives") %>%
  mutate(nonneg_ok = n_negatives == 0) %>%
  arrange(nonneg_ok, desc(n_negatives))

summary_nonneg
```

Great. No negative counts slipped their way into the dataset somehow.


# Subsetting the Data

The [Unified Rules of Mixed Martial Arts](https://www.ufc.com/unified-rules-mixed-martial-arts) was adopted November 2000. So let's start from there. Our data will include fights starting from UFC 28: High Stakes.

```{r}
unified_rules_adopted = "2000-11-01" # YYYY-MM-DD
```


```{r}
fights <- fights %>%
  filter(date >= as.Date(unified_rules_adopted))
```


How many rows have less than 5 significant strikes thrown?

```{r}
sum(fights$sig_strikes_thrown < 5, na.rm = TRUE)
```

What percent of the data is that?

```{r}
mean(fights$sig_strikes_thrown < 5, na.rm = TRUE)
```

4% of fights have fighters who threw less than 5 significant strikes.


What about just the plain old strikes thrown?

```{r}
sum(fights$strikes_thrown < 5, na.rm = TRUE)
```


```{r}
mean(fights$strikes_thrown < 5, na.rm = TRUE)
```

Around 3% of fights have fighters who threw less than 5 total strikes.


Good to know.

# Data Summaries

Get an overall feel for the numeric data columns.

```{r}
fights %>%
  select(where(is.numeric)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(
    n = sum(!is.na(value)),
    # missing = sum(is.na(value)), # already verified nothing missing
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    q25 = quantile(value, 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    q75 = quantile(value, 0.75, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    .groups = "drop"
  )
```

Many of these columns have the same distributions due to correlations and the reciprocal nature of fights: two fighters, the `sig_strikes_landed` for fighter 1 in one row is going to be the `sig_strikes_landed_by_opp` in fighter 2's row for their fight (`fight_id`).


# Preliminary Plots

```{r}
library(ggplot2)

theme_striking <- function(base_size = 11, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      # Titles
      plot.title = element_text(
        face = "bold",
        hjust = 0,
        size = base_size + 2
      ),
      plot.subtitle = element_text(
        hjust = 0,
        size = base_size
      ),
      plot.caption = element_text(
        hjust = 1,
        size  = base_size - 2,
        color = "grey40"
      ),
      
      # Axes
      axis.title.x = element_text(
        margin = margin(t = 8),
        face   = "bold"
      ),
      axis.title.y = element_text(
        margin = margin(r = 8),
        face   = "bold"
      ),
      axis.text.x = element_text(
        size  = base_size - 1,
        angle = 0,
        vjust = 1
      ),
      axis.text.y = element_text(
        size = base_size - 1
      ),
      
      # Grid lines: a bit lighter since striking plots can be busy
      panel.grid.major.x = element_line(linewidth = 0.25),
      panel.grid.major.y = element_line(linewidth = 0.25),
      panel.grid.minor   = element_blank(),
      
      # Facet strips
      strip.background = element_rect(
        fill  = "grey90",
        color = "grey70",
        linewidth = 0.5
      ),
      strip.text = element_text(
        face = "bold",
        size = base_size - 1
      ),
      
      # Legend
      legend.position = "right",
      legend.title    = element_text(face = "bold"),
      legend.text     = element_text(size = base_size - 1),
      
      # Margins
      plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
    )
}
```



```{r}
# Strikes landed vs. strikes thrown - all fighters across all their fights
ggplot(fights, aes(x = strikes_thrown, y = strikes_landed)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype="dashed") +
  coord_equal() +
  labs(
    x = "Strikes Thrown",
    y = "Strikes Landed",
    title = "Strikes Landed vs. Strikes Thrown"
  ) +
  theme_striking()
```



```{r}
# Strikes landed vs. strikes thrown
sig_strikes_landed_vs_thrown_scatter <- ggplot(fights, aes(x = sig_strikes_thrown, y = sig_strikes_landed)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype="dashed", color = "chartreuse4", linewidth = 0.7) +
  coord_equal() +
  labs(
    x = "Significant Strikes Thrown",
    y = "Significant Strikes Landed",
    title = "Significant Strikes Landed vs. Significant Strikes Thrown"
  ) +
  theme_striking()

sig_strikes_landed_vs_thrown_scatter
```

```{r}
ggsave(
  filename = here("reports", "midterm", "figs", "sig_strikes_landed_vs_thrown_scatter.png"),
  plot     = sig_strikes_landed_vs_thrown_scatter,
  width    = 6,
  height   = 4,
  dpi      = 300,
  bg       = "white"
)
```



Strikes and significant strikes are correlated for sure. In particular, *significant* strikes are a subset of strikes. 
- What counts as a "significant strike" in the UFC?
- How is it different different from just a regular strike?

These definitions are inherently subjective.


```{r}
# Strikes landed vs. strikes thrown
ggplot(fights, aes(x = sig_strikes_thrown, y = sig_strikes_landed, color = weight_class)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype="dashed") +
  labs(
    x = "Significant Strikes Thrown",
    y = "Significant Strikes Landed",
    title = "Significant Strikes: Landed vs. Thrown by Weight Class"
  ) +
  theme_striking()
```

Looks like a mess. Too many colors.


Visualize distribution of significant strike accuracy (for fights with more than 5 significant strikes thrown).

```{r}
# Include only fights where the fighter (pertaining to the row) threw more than 5 significant strikes
MIN_SIG_STRIKES_THROWN = 5

sig_strike_acc_hist <- fights %>%
  filter(sig_strikes_thrown >= MIN_SIG_STRIKES_THROWN) %>%
  mutate(sig_strike_acc = sig_strikes_landed / sig_strikes_thrown) %>%
  ggplot(aes(x = sig_strike_acc)) +
  geom_histogram(binwidth = 0.05, fill = "firebrick", color = "white") +
  labs(
    title = "Distribution of Signficant Strike Accuracy",
    x = "Significant Strike Accuracy",
    y = "Count"
  ) +
  theme_striking()

sig_strike_acc_hist
```

```{r}
ggsave(
  filename = here("reports", "midterm", "figs", "sig_strike_acc_hist.png"),
  plot     = sig_strike_acc_hist,
  width    = 6,
  height   = 4,
  dpi      = 300,
  bg       = "white"
)
```



```{r}
# Include only fights where the fighter (pertaining to the row) threw more than 5 significant strikes
MIN_SIG_STRIKES_THROWN = 5

fights_2 <- fights %>% 
  filter(sig_strikes_thrown > MIN_SIG_STRIKES_THROWN) %>%
  mutate(sig_strike_acc = sig_strikes_landed / sig_strikes_thrown)
```


```{r}
levels_men <- c(
  "Flyweight","Bantamweight","Featherweight","Lightweight",
  "Welterweight","Middleweight","Light Heavyweight","Heavyweight","Catchweight"
)
```


```{r}
sig_strike_acc_by_weight_men <- fights_2 %>%
  filter(gender == "Men") %>%
  mutate(weight_class = forcats::fct_relevel(weight_class, levels_men)) %>%
  ggplot(aes(x = sig_strike_acc, fill = weight_class)) +
  geom_histogram(binwidth = 0.05, color = "white", show.legend = FALSE) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~ weight_class, ncol = 3, drop = TRUE) +
  labs(
    title = "Men's Distribution of Signficant Strike Accuracy Across Weight Classes",
    x = "Significant Strike Accuracy",
    y = "Count"
  ) +
  theme_striking()

sig_strike_acc_by_weight_men
```


```{r}
ggsave(
  filename = here("reports", "midterm", "figs", "sig_strike_acc_by_weight_men.png"),
  plot     = sig_strike_acc_by_weight_men,
  width    = 8,
  height   = 5,
  dpi      = 300,
  bg       = "white"
)
```



```{r}
levels_women <- c(
  "Strawweight", "Flyweight", "Bantamweight", "Featherweight"
)
```


```{r}
sig_strike_acc_by_weight_women <- fights_2 %>%
  filter(gender == "Women") %>%
  mutate(weight_class = forcats::fct_relevel(weight_class, levels_women)) %>%
  ggplot(aes(x = sig_strike_acc, fill = weight_class)) +
  geom_histogram(binwidth = 0.05, color = "white", show.legend = FALSE) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~ weight_class, ncol = 2, drop = TRUE) +
  labs(
    title = "Women's Distribution of Signficant Strike Accuracy Across Weight Classes",
    x = "Significant Strike Accuracy",
    y = "Count"
  ) +
  theme_striking()

sig_strike_acc_by_weight_women
```


```{r}
ggsave(
  filename = here("reports", "midterm", "figs", "sig_strike_acc_by_weight_women.png"),
  plot     = sig_strike_acc_by_weight_women,
  width    = 8,
  height   = 5,
  dpi      = 300,
  bg       = "white"
)
```


Distribution of significant strikes landed vs. significant strikes avoided.

```{r}
sig_strikes_landed_vs_avoided_scatter <- fights_2 %>%
  ggplot(aes(x = sig_strikes_landed, y = sig_strikes_avoided)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype="dashed", color = "chartreuse4", linewidth = 0.7) +
  coord_equal() +
  labs(
    x = "Significant Strikes Landed",
    y = "Significant Strikes Avoided",
    title = "Significant Strikes: Landed vs. Avoided"
  ) +
  theme_striking()

sig_strikes_landed_vs_avoided_scatter
```


```{r}
ggsave(
  filename = here("reports", "midterm", "figs", "sig_strikes_landed_vs_avoided_scatter.png"),
  plot     = sig_strikes_landed_vs_avoided_scatter,
  width    = 8,
  height   = 5,
  dpi      = 300,
  bg       = "white"
)
```


From cursory inspection, it looks like quite a lot more significant strikes are avoided than are landed.



How does significant strike accuracy vary across fighters? What is the mean significant strike accuracy for each of the fighters? What does that distribution look like?
- Here I'm looking at fighters who have fought at least 3 times across their career (so far).

```{r}
MIN_NUM_FIGHTS = 3

per_fighter <- fights_2 %>%
  group_by(fighter_id, fighter) %>%
  summarize(
    mean_sig_acc = mean(sig_strike_acc, na.rm = TRUE),
    num_fights = n(),
    .groups = "drop"
  )

per_fighter %>%
  filter(num_fights >= MIN_NUM_FIGHTS) %>%
  filter(is.finite(mean_sig_acc), between(mean_sig_acc, 0, 1)) %>%
  ggplot(aes(x = mean_sig_acc)) +
  geom_histogram(binwidth = 0.05, color = "white") +
  scale_x_continuous(labels = scales::percent_format(1)) +
  labs(title = "Per-Fighter Mean Significant Strike Accuracy",
       x = "Mean Accuracy", y = "Count") +
  theme_striking()
```

This is the average of fight significant strike accuracies.
- Compute significant strike accuracy per fight, then average across a fighter's fights.
- Every fight gets equal weight, no matter how many significant strikes were thrown.



```{r}
per_fighter_weighted <- fights_2 %>%
  group_by(fighter_id, fighter) %>%
  summarize(
    total_landed = sum(sig_strikes_landed, na.rm = TRUE),
    total_thrown = sum(sig_strikes_thrown, na.rm = TRUE),
    overall_sig_acc = if_else(total_thrown > 0, total_landed / total_thrown, NA_real_),
    num_fights = n(),
    .groups = "drop"
  )

per_fighter_weighted %>%
  filter(num_fights >= MIN_NUM_FIGHTS) %>%
  filter(is.finite(overall_sig_acc), between(overall_sig_acc, 0, 1)) %>%
  ggplot(aes(x = overall_sig_acc)) +
  geom_histogram(binwidth = 0.05, color = "white") +
  scale_x_continuous(labels = scales::percent_format(1)) +
  labs(title = "Per-Fighter Overall Significant Strike Accuracy",
       x = "Overall accuracy", y = "Count") +
  theme_striking()

```

This is overall accuracy. The number of significant strikes landed across all fights of a fighter divided by the number of significant strikes thrown across all fights by said fighter.
- Pool all landed and all thrown, then take one ratio.
- This number is weighted by the number of significant strikes thrown.


The basic accuracy metric of each fighter, for each fight, will be the ratio between the amount of strikes (punches, kicks, knees, elbows) that hit the opponent and the total number of strikes they threw.

$$\text{striking accuracy} = \frac{\text{strikes landed}}{\text{strikes thrown}}$$
Import caveat: What counts as a strike and how does the UFC count a strike?
- There's a lot of subjectivity to this.
- The way it works it that there's probably a few people at these live events whose job is to count the number of strikes thrown and landed. Then they review the footage afterward and correct their tallies, maybe? Not entirely sure. Need to read up on that.
- What is the difference between a strike and a significant strike? What makes a strike significant? Need to understand that distinction.



All fighters, across all their fights, are never 100% accurate with their strikes. They always throw more strikes than they land. Captain Obvious over here, but good to know we can see it in the data.

# Career (Survival Analysis)

```{r}
library(dplyr)
library(lubridate)
```

```{r}
today <- as_date(Sys.Date())         # or fix a reference date
inactive_cutoff_years <- 2

careers <- fights %>%
  group_by(fighter_id) %>%
  summarise(
    fighter = first(fighter),
    gender = first(gender),
    weight_class = first(weight_class),
    first_fight = min(date),
    last_fight = max(date),
    n_fights = n()
  ) %>%
  mutate(
    career_length_years = as.numeric(last_fight - first_fight) / 365,
    inactive_years = as.numeric(today - last_fight) / 365,
    event = if_else(inactive_years >= inactive_cutoff_years, 1, 0)
  )
```

If `event` is 1, then the fighter has been inactive for at least 2 years. Otherwise, if `event` is 0, then the fighter has been inactive for less than that. We treat this as still active, so the observation is right censored.


```{r}
careers %>%
  count(event)
```

```{r}
library(ggplot2)

theme_careers <- function(base_size = 12, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      # Titles and subtitles
      plot.title = element_text(
        face = "bold",
        hjust = 0,
        size = base_size + 2
      ),
      plot.subtitle = element_text(
        hjust = 0,
        size = base_size
      ),
      
      # Axis titles and text
      axis.title.x = element_text(
        margin = margin(t = 8),
        face = "bold"
      ),
      axis.title.y = element_text(
        margin = margin(r = 8),
        face = "bold"
      ),
      axis.text = element_text(size = base_size - 1),
      
      # Panel grid lines
      panel.grid.major.x = element_line(linewidth = 0.3),
      panel.grid.major.y = element_line(linewidth = 0.3),
      panel.grid.minor = element_blank(),
      
      # Legend
      legend.position = "bottom",
      legend.title = element_blank(),
      legend.text = element_text(size = base_size - 1),
      
      # Plot margins
      plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
    )
}

```



```{r}
career_length_hist <- ggplot(careers, aes(career_length_years)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of UFC Career Lengths",
    x = "Career Length (years)",
    y = "Number of Fighters"
  ) +
  theme_careers()

career_length_hist
```

```{r}
ggsave(
  filename = here("reports", "midterm", "figs", "career_length_hist.png"),
  plot     = career_length_hist,
  width    = 8,
  height   = 5,
  dpi      = 300,
  bg       = "white"
)
```




```{r}
career_length_by_weight_boxplot <- ggplot(careers, aes(x = weight_class, y = career_length_years, fill = weight_class)) +
  geom_boxplot(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Career Length by Weight Class",
    x = "Weight Class",
    y = "Career Length (years)"
  ) +
  scale_fill_brewer(palette = "Set3") +
  theme_careers()

career_length_by_weight_boxplot
```


```{r}
ggsave(
  filename = here("reports", "midterm", "figs", "career_length_by_weight_boxplot.png"),
  plot     = career_length_by_weight_boxplot,
  width    = 8,
  height   = 5,
  dpi      = 300,
  bg       = "white"
)
```



```{r}
ggplot(careers, aes(n_fights, career_length_years)) +
  geom_point(alpha = 0.3) +
  labs(
    title = "Number of UFC Fights vs Career Length",
    x = "Total UFC Fights",
    y = "Career Length (years)"
  ) +
  theme_careers()
```



```{r}
library(survival)
library(survminer)

km <- survfit(Surv(career_length_years, event) ~ 1, data = careers)

ggsurvplot(
  km,
  conf.int = TRUE,
  risk.table = TRUE,
  xlab = "Career length (years)",
  ylab = "Proportion still active",
  title = "Survival Curve for UFC Careers"
)
```



# Model-Ready Data


Since we're building a model to estimate the striking ability of fighters, we need to drop rows where there were no significant strikes thrown. We use the ratio between the significant strikes landed and thrown as noisy-but-good-enough measures of each fighter's offensive ability in a fight. Similarly, we use the ratio between the significant strikes landed and thrown by the opponent as a measure of each fighter's defensive ability.





























































